---
title: "Module-Five"
output: html_document
date: "2025-02-03"
---

read.table() - reads and pulls in the data from a data file
read.csv() 
file.choose() - allows you to click which file you want
head() - lists first 6 lines of data by default, can read more by using brackets. Ex: head(data)[1:10]
tail() - lists last 6 lines of data by default

```{r}
file <- file.choose()
##chose the CPDS reduced txt file
data <- read.table(file, header = TRUE, sep = "\t", stringsAsFactors = FALSE, fill = T)
head(data)
```
You can also get a file by writing the pathway to it, as demonstrated below with a file that is in the desktop folder.

**readr package**
Offers additional functions to read delimited (separated, for example by commas or spaces) text files

To access these functions, use command require(readr)

**Loading Excel Files**
There are several packages available that provide functions available for loading data into R from Excel.
{readxl} and {XLConnect} are two of the fastest and easiest-to-use

To use these packages use the require() command.

```{r}
require(readxl)
file <- "~/Desktop/CPDS-1960-2014-reduced.xlsx"
data <- read_excel(file, sheet = 1, col_names = TRUE)
head(data)
```

**Importing from a remote server (such as github)**

First copy the url of the raw data you are trying to pull in.

If using {curl} use the command library(curl) first.

For a comma-separated value text file (a .csv file)...
```{r}
library(curl)
file <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/refs/heads/master/AN588_Fall23/CPDS-1960-2014-reduced.csv")
data <- read.csv(file, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(data)
```
For a tab-delimited file (.tsv or .txt file)...
```{r}
library(curl)
file <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/refs/heads/master/AN588_Fall23/CPDS-1960-2014-reduced.txt")
data <- read.table(file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
head(data)
```

If using {readr}, use the command library(readr).

txt file example...
```{r}
library(readr)
file <- "https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/refs/heads/master/AN588_Fall23/CPDS-1960-2014-reduced.txt"
data <- read.table(file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
head(data)
```
